% =============================================================================
% ABSTRACT
% =============================================================================

\begin{abstract}

This thesis presents a comprehensive study of Higher Order Spectra (HOS) analysis techniques for fault detection in rotordynamic systems. The research focuses on the application of advanced spectral analysis methods, including bispectrum and trispectrum analysis, combined with machine learning algorithms for automated fault classification in rotating machinery.

The study begins with a thorough literature review of spectral analysis techniques and their applications in rotordynamics. The theoretical foundation covers the mathematical principles of higher-order statistics, Fourier analysis, and signal processing methods relevant to vibration analysis. A systematic methodology is developed for feature extraction from vibration signals, incorporating both traditional spectral features and higher-order statistical measures.

The implementation includes a comprehensive Python-based framework for spectral analysis, featuring modules for signal processing, feature extraction, and machine learning classification. The system is validated using both simulated and real-world vibration data from rotating machinery, including bearing fault datasets and rotor unbalance scenarios.

Experimental results demonstrate the effectiveness of higher-order spectral features in distinguishing between different fault types, with particular success in detecting bearing defects, rotor unbalance, and misalignment conditions. Machine learning classifiers, including Random Forest, Support Vector Machines, and Neural Networks, are evaluated and compared for their performance in fault classification tasks.

The research contributes to the field by providing a systematic approach to combining higher-order spectral analysis with modern machine learning techniques for improved fault detection accuracy. The developed framework offers practical tools for condition monitoring applications in industrial rotating machinery.

\textbf{Keywords:} Higher Order Spectra, Rotordynamics, Fault Detection, Machine Learning, Spectral Analysis, Vibration Analysis, Condition Monitoring

\end{abstract}
